Intrinsic Verification of Parsers and Formal Grammar Theory in Dependent Lambek Calculus

### 2024-11-08 paper/paper.tex
Turing machines

\subsection{Turing Machines}
\newcommand{\Turing}{\mathtt{Turing}}
\newcommand{\ttt}{\mathtt{t}}
\newcommand{\T}{\mathtt{T}}
\newcommand{\acc}{\mathtt{acc}}
\newcommand{\rej}{\mathtt{rej}}
\newcommand{\Tape}{\mathtt{Tape}}
\newcommand{\parseTape}{\mathtt{parseTape}}
\newcommand{\snoc}{\mathtt{snoc}}
\steven{This subsection is still a WIP}
%% 5. Expressiveness: Turing Machines
The expressive power of \theoryabbv spans the entire Chomsky hierarchy, as we may
encode the accepting traces of Turing machines as linear types.

\steven{TODO introduce tape alphabet and blank symbol}

The data of a Turing machine $\T$  is given by the following data,
\begin{itemize}
  \item $\T.\states$ --- a finite set of states
  \item $\T.\init : \T.\states$ --- the initial state
  \item $\T.\acc , \T.\rej : \T.\states$ --- distinguished accepting and
        rejecting states, respectively, such that $\T.\acc \neq \T.\rej$
  \item $\T.\delta : \T.\states \to \Gamma \to \T.\states \times \Gamma \times \Bool$
        --- a deterministic transition function. Given the current state and the
        head of the tape return the new state, the character written to the
        tape, and which direction to shift ($\true$ for left, $\false$ for right)
\end{itemize}

The tape of a
(one-sided) Turing machine is encoded as nonlinear function from the natural numbers to the tape
alphabet $\Gamma$, $\Tape = \mathbb{N} \to \Gamma$. To initialize the tape of
$\T$, we define the parse transformer
$\parseTape : \ltonl {(\stringg \lto \LinSigTy {\ttt} {\Tape} {\Turing~\ttt})}$ in \cref{fig:parseTape}.

\steven{TODO finish the description of Turing machines}

\begin{figure}
\begin{lstlisting}
data Turing : (t : Tape) (*@$\to$@*) L where
  nil : (*@$\uparrow$@*)(&[ t (*@$\in$@*) Tape ] Turing t)
  snoc : (*@$\uparrow$@*)(&[ c (*@$\in$@*) (*@$\Sigma$@*) ] &[ t (*@$\in$@*) Tape ](Turing t (*@$\lto$@*) c (*@$\lto$@*) Turing (push c t)))
\end{lstlisting}
\caption{Inductive linear type of Turing machines initialized with tape $\ttt$}
\label{fig:turinginductive}
\end{figure}

\begin{figure}
\begin{lstlisting}
parseTape : (*@$\uparrow$@*)(String (*@$\lto$@*) (*@$\oplus$@*)[ t (*@$\in$@*) Tape ] Turing t)
parseTape String.nil = (*@$\sigma$@*) ((*@$\lambda$@*) n . blank) (Turing.nil ((*@$\lambda$@*) n . blank))
parseTape (String.cons ((*@$\sigma$@*) c a) w) =
  let (*@$\sigma$@*) t Tur = parseTape w in (*@$\sigma$@*) (push c t) (snoc c t Tur a)
\end{lstlisting}
\caption{Initialization of the tape of Turing machine as a parse transformer}
\label{fig:parseTape}
\end{figure}

We may then define a term of type
$\mathsf{parseTape} : \ltonl {\left( \stringg \lto \LinSigTy {\ttt} {\Tape} {\Turing~\ttt} \right)}$.
In this manner, all of the interesting computations that the Turing machine
would do occur solely inside of the nonlinear index. That is,
$\mathsf{parseTape}$ does not discern which strings are accepted by the Turing
machine, nor does it do any computation. $\mathsf{parseTape}$ simply loads the
input string into the initial spaces on the tape.

Once the string has been loaded onto the tape, we may further define the
following term to run the Turing machine.
\[
  \mathsf{run} : \ltonl {\left( \bigoplus\limits_{t : \mathbb{N} \to \Gamma} \Turing(t) \lto \bigwith\limits_{\mathsf{fuel} : \mathbb{N}} \mathsf{Maybe} \left( \bigoplus\limits_{t : \mathbb{N} \to \Gamma} \bigoplus\limits_{b : \Bool} \bigoplus\limits_{\mathsf{trace} : \mathsf{TuringTrace}(b , t)} \Turing(t) \right) \right)}
\]

where $\mathsf{TuringTrace}$ is a nonlinear type denoting terminating traces
through the Turing machine. Note that $\mathsf{run}$ returns a $\mathsf{Maybe}$
type in a fuel parameter. Turing machines may not terminate. If they do
terminate within $\mathsf{fuel}$-many steps, then we return the trace along with
the current tape and a boolean denoting if the trace is accepting or rejecting.
If the trace does not terminate within the allotted amount of fuel, then we
timeout and return nothing.

>>>>

\subsection{Turing Machines}
\newcommand{\Turing}{\mathtt{Turing}}
\newcommand{\ttt}{\mathtt{t}}
\newcommand{\blank}{\mathtt{blank}}
\newcommand{\push}{\mathtt{push}}
\newcommand{\T}{\mathtt{T}}
\newcommand{\acc}{\mathtt{acc}}
\newcommand{\rej}{\mathtt{rej}}
\newcommand{\Tape}{\mathtt{Tape}}
\newcommand{\parseTape}{\mathtt{parseTape}}
\newcommand{\snoc}{\mathtt{snoc}}
%% 5. Expressiveness: Turing Machines
The expressive power of \theoryabbv spans the entire Chomsky hierarchy, as we may
encode the accepting traces of Turing machines as linear types.

The data of a Turing machine $\T$  is given by the following data,
\begin{itemize}
  \item $\Gamma$ --- the tape alphabet, a superset of the input alphabet $\Sigma$
  \item $\blank : \Gamma$ --- a distinguished character that marks an empty cell on the tape
  \item $\T.\states$ --- a finite set of states
  \item $\T.\init : \T.\states$ --- the initial state
  \item $\T.\acc , \T.\rej : \T.\states$ --- distinguished accepting and
        rejecting states, respectively, such that $\T.\acc \neq \T.\rej$
  \item $\T.\delta : \T.\states \times \Tape \times \mathbb{N} \to \T.\states \times \Tape \times \mathbb{N}$
        --- a deterministic transition function. $\T.\delta$ reads the current state, the
        current tape, and the index of the head of the tape and then returns the
        updated state, tape, and head.
\end{itemize}

\begin{figure}
\begin{lstlisting}
data Turing(*@$_{\T}$@*) : (t : Tape) (*@$\to$@*) L where
  nil : (*@$\uparrow$@*)(&[ t (*@$\in$@*) Tape ] Turing(*@$_{\T}$@*) t)
  snoc : (*@$\uparrow$@*)(&[ c (*@$\in$@*) (*@$\Sigma$@*) ] &[ t (*@$\in$@*) Tape ](Turing t (*@$\lto$@*) c (*@$\lto$@*) Turing(*@$_{\T}$@*) (push c t)))
\end{lstlisting}
\caption{Inductive linear type of Turing machines initialized with tape $\ttt$}
\label{fig:turinginductive}
\end{figure}

\begin{figure}
\begin{lstlisting}
parseTape : (*@$\uparrow$@*)(String (*@$\lto$@*) (*@$\oplus$@*)[ t (*@$\in$@*) Tape ] Turing(*@$_{\T}$@*) t)
parseTape String.nil = (*@$\sigma$@*) ((*@$\lambda$@*) n . blank) (Turing.nil ((*@$\lambda$@*) n . blank))
parseTape (String.cons ((*@$\sigma$@*) c a) w) =
  let (*@$\sigma$@*) t Tur = parseTape w in (*@$\sigma$@*) (push c t) (snoc c t Tur a)
\end{lstlisting}
\caption{Initialization of the tape of Turing machine as a parse transformer}
\label{fig:parseTape}
\end{figure}

The tape of a
(one-sided) Turing machine is encoded as nonlinear function from the natural numbers to the tape
alphabet $\Gamma$, $\Tape = \mathbb{N} \to \Gamma$.
We define $\push~\c~\ttt : \Tape \to \Tape$ to increment the index of every
character in the tape and insert $\c$ at index $0$. To run $\T$ on
the input string $\w$, we write $\w$ to the leftmost entries of the tape and
then repeatedly transition according to $\T.\delta$. If the machine enters the
states $\T.\acc$ or $\T.\rej$, then the execution of $\T$ halts and either
accepts or rejects accordingly.

In \cref{fig:turinginductive}, we give an inductive linear type that describes
Turing machines initialized with a given tape. However, unlike the types of
traces for finite automata, the execution of a Turing machine is not described
directly by the linear type system. We simply use the type $\Turing_{\T}$ to encode
the initial state of the machine. To write the input string to the leftmost
entries of the tape, we define the parse transformer
$\parseTape : \ltonl {(\stringg \lto \LinSigTy {\ttt} {\Tape} {\Turing_{\T}~\ttt})}$
in \cref{fig:parseTape}. Once the tape has been initialized, we capture the
execution of $\T$ via the non-linear types.
% data TuringTrace (b : Bool) : ⟨ Q ⟩ × Tape × Head → Type ℓ-zero where
%   accept : ∀ t h → b ≡ true → TuringTrace b (acc , t , h)
%   reject : ∀ t h → b ≡ false → TuringTrace b (rej , t , h)
%   move : ∀ q t h →
%     TuringTrace b (transition q t h) →
%     TuringTrace b (q , t , h)
\begin{figure}
\begin{lstlisting}
data Trace(*@$_{\T}$@*) : Bool (*@$\to$@*) T.states (*@$\times$@*) Tape (*@$\times$@*) (*@$\mathbb{N}$@*) (*@$\to$@*) U where
  accept : (*@$\forall$@*) (t : Tape) (n : (*@$\mathbb{N}$@*)) (*@$\to$@*) Trace(*@$_{\T}$@*) true (T.acc , t , n)
  reject : (*@$\forall$@*) (t : Tape) (n : (*@$\mathbb{N}$@*)) (*@$\to$@*) Trace(*@$_{\T}$@*) false (T.rej , t , n)
  move : (*@$\forall$@*) (b : Bool) (s : T.states) (t : Tape) (n : (*@$\mathbb{N}$@*)) (*@$\to$@*)
             Trace(*@$_{\T}$@*) b (T.(*@$\delta$@*) s t n) (*@$\to$@*) Trace(*@$_{\T}$@*) b (s , t , n)
\end{lstlisting}
\caption{Traces through a Turing machine as a non-linear inductive type}
\label{fig:turingtrace}
\end{figure}

In \cref{fig:turingtrace} we define $\Trace_{\T}$ as a nonlinear type indexed by
an acceptance Boolean and start configuration. A configuration of a Turing
machine is the data of the start state, start tape, and start index for the head
of tape. The
traces of $\T$ either accept, reject, or transitions as described by $\T.\delta$.

Once the input string has been loaded onto the tape, we execute $\T$ by defining
a Turing machine interpreter of the following type,

\[
  \texttt{run} : \ltonl {\left( \bigoplus\limits_{\ttt : \Tape} \Turing_{\T}~\ttt \lto \bigwith\limits_{\texttt{fuel} : \mathbb{N}} \texttt{Maybe} \left(  \bigoplus\limits_{\b : \Bool} \bigoplus\limits_{\s : \T.\states}  \bigoplus\limits_{\texttt{n} : \mathbb{N}} \bigoplus\limits_{\ttt : \Tape}  \bigoplus\limits_{\texttt{tr} : \Trace_{\T}~\b~(\s , \ttt , \texttt{n})} \Turing_{\T}~\ttt \right) \right)}
\]

$\mathtt{run}$ returns a result of $\mathtt{Maybe}$
type varying over a fuel\footnote{Putting termination issues aside, we may omit
  this fuel parameter. Although it is necessary when defining a Turing machine
  interpreter in a total language like Agda.} parameter. It takes transitions at most
$\texttt{fuel}$-many transitions in $\T$, and it returns a halting trace if one is
found or timeouts otherwise. The data of this trace is encoded in the nonlinear
index $\texttt{tr} : \Trace_{\T}~\b~(\s , \ttt , \texttt{n})$. Importantly, when
$\texttt{tr}$ is an \emph{accepting} trace then the grammar
\[
\bigoplus\limits_{\s : \T.\states}  \bigoplus\limits_{\texttt{n} : \mathbb{N}} \bigoplus\limits_{\ttt : \Tape}  \bigoplus\limits_{\texttt{tr} : \Trace_{\T}~\true~(\s , \ttt , \texttt{n})} \Turing_{\T}~\ttt
\]
describes the strings that are accepted by $\T$. Therefore, we may encode
arbitrary recusively enumerable languages within \theoryabbv.
